  document.RMD template - Version 1 - 5/6/2024 
  
  Note: to replace the current rstudio templatejust rename your current "document.RMD" and copy this one into "C:\Program Files\RStudio\resources\app\resources\templates" or your install directory 
  
  - feel free to add in any other helpful formatting, functions etc.
  - also delete this template header



# Analysis overview 



# setup and environment including helpful function fs() to "forwardslash" paths on your clipboard

```{r}

#helpful environment modifications

  #(optional) set number of columns to display in "View()" function to 1000 - revert to 50 if you don't like the behavior
  if(F) rstudioapi::writeRStudioPreference("data_viewer_max_columns", 1000L)

  #turn off scientific notation 
  options(scipen = 10E6)
  
  #prevent strings being read as factors
  options(stringsAsFactors = F)

# helpful functions

  #convenience function for more usable version of writeClipboard - overcomes newline issues
  writeClipboard2 = function(x) cat(x, file="clipboard")
  
  #generic file path editing function - forwards or backwards
  reslash<- function(x=NA, slash=c("back","forward")){
    if(is.na(x[1])) x = readClipboard()
    if(slash[1]=="back") path <- shQuote(gsub("/","\\\\",gsub("\\", "\\\\", x, fixed = TRUE), fixed = TRUE))
    if(slash[1]=="forward") path <- shQuote(gsub("\\", "/", x, fixed = TRUE))
    writeClipboard2(path)
    return(path)
  }

  #correctly escape back slashes and quote path - a parameterized version o f   
  bs=function(x=NA){reslash(x,slash="back")}
  
  #correctly escape back slashes and quote path   
  fs=function(x=NA){reslash(x,slash="forward")}
  
  #function to grab names from data.frame, quote them, place commas between them, and send to clipboard
  nmsVec=function(x){x=paste("c('",paste(names(x),collapse="','"),"')",sep=""); writeClipboard2(x);return(x)}
  
  #function to grab names from data.frame, quote them, place commas between them, and send to clipboard
  # - version 2 adds newline at 50 chars
  nmsVec2=function(x,char_break=70,newline="\n"){
    nms_in = names(x)
    vec_i = paste0("'",nms_in[1],"'")
    char_break_in = char_break
    for(i in 2:length(nms_in)){
      vec_ln = nchar(vec_i)
      
      if((vec_ln + nchar(nms_in[i])) >= char_break_in){
        char_break_in = char_break + nchar(vec_i) 
        vec_i = paste0(vec_i,"\n    ,'",nms_in[i],"'")
      }else{
        vec_i = paste0(vec_i,",'",nms_in[i],"'")
      }
    }
    #add parentheses  
    vec_i = paste("c(",vec_i,"  \n    )",sep="",collapse="")
    
    writeClipboard2(vec_i)
    return(vec_i)
  }


  #mse and rmse functions
  mse = function(err,na.rm=T){mean(err^2,na.rm=na.rm)} 
  rmse = function(err,na.rm=T){sqrt(mean(err^2,na.rm=na.rm))}   
  
  #file_tif = RForInvt::file_increment(file.path("c:/temp","some_plots.tif"), increment=T)
  #con_this = DBI::dbConnect(RSQLite::SQLite(), file_db )
  #tiff(file_tif , width = 6, height = 4, units = "in", res = 600)
  #dev.off()
  
  
```

## load data - some examples of ways to load various data types. 

```{r}

#examples
if(F){ 
	
  library(RSQLite)
  library(DBI)
	
	#Example make some fake data and write to versioned files 
  
  	#create new output file paths for this version - increment = T  
    dir_temp = "c:/temp"
  	file_csv = RForInvt::file_increment(file.path(dir_temp,"some_data.csv"), increment=T)
  	file_rds = RForInvt::file_increment(file.path(dir_temp,"some_data.RDS"), increment=T)
  	file_db = RForInvt::file_increment(file.path(dir_temp,"some_data.sqlite"), increment=T)
  	
  	#simulate / write small simulated table of data
  	
  		#simulate
  		test_dat = data.frame(plot=1:5,some_y=rnorm(5))
  		
  		#write csv
  		write.csv(test_dat,  file_csv )
  		
  		#save R binary format
  		saveRDS(test_dat, file_rds)
  		
  		#save to database
  	  con_this = dbConnect(RSQLite::SQLite(), file_db )
  	  	dbWriteTable( con_this , "pldat" , test_dat , overwrite = T )
  	  dbDisconnect(con_this)  
	
	#grab latest versions of data - increment = F
	  
	  #prep file names
    dir_temp = "c:/temp"
  	file_csv = RForInvt::file_increment(file.path(dir_temp,"some_data.csv"), increment=F)
  	file_rds = RForInvt::file_increment(file.path(dir_temp,"some_data.RDS"), increment=F)
  	file_db = RForInvt::file_increment(file.path(dir_temp,"some_data.sqlite"), increment=F)
  	
  	#read data csv/rds
  	dat_csv = read.csv(file_csv)
  	dat_rds = readRDS(file_rds)
  	
  	#read sqlite data
    con_this = dbConnect(RSQLite::SQLite(), file_db )
    	dat_db = dbReadTable(con_this ,"pldat" )
    dbDisconnect(con_this)  
    
    #add sf::st_write example
    
    #add terra::writeRaster example
    
}


```

## example of how to save versions figures to tiff and pdf

```{r, eval=FALSE}

	#create versioned files names
  dir_temp = "c:/temp"
	file_tif = RForInvt::file_increment(file.path(dir_temp,"some_plots.tif"), increment=T)
	file_pdf = RForInvt::file_increment(file.path(dir_temp,"some_plots.pdf"), increment=T)

	#simulate / write small simulated table of data
	
		#simulate
		test_dat1 = data.frame( plot=1:5,y=rnorm(5), x=rnorm(5) )
		
		#save tif - show how to combine exponent with regular text
		tiff(file_tif , width = 6, height = 4, units = "in", res = 600)
		  plot(y~x, data = test_dat1 , main = "test_dat1", xlab=expression(paste("bigfoot height ",cm^2)), ylab="ft")
		dev.off()
		
		#save pdf - show how to save 2 figures in same file
		pdf(file_pdf , width = 6, height = 4, units = "in", res = 600)
		  plot(y~x, data = test_dat1 , main = "test_dat1", xlab=expression(paste("bigfoot height ",cm^2)), ylab="ft")
		  plot(y~x, data = test_dat1 , main = "test_dat1", xlab=expression(paste("bigfoot height ",cm^2)), ylab="ft")
		dev.off()



```

## 

```{r, eval=FALSE}

```

## 

```{r, eval=FALSE}

```

## 

```{r, eval=FALSE}

```


## 

```{r, eval=FALSE}

```

## 

```{r, eval=FALSE}

```

## 

```{r, eval=FALSE}

```
