  document.RMD template - Version 1 - 5/6/2024 
  
  Note: to replace the current rstudio templatejust rename your current "document.RMD" and copy this one into "C:\Program Files\RStudio\resources\app\resources\templates" or your install directory 
  
  - feel free to add in any other helpful formatting, functions etc.
  - also delete this template header



# Analysis overview 


# setup and environment including helpful function fs() to "forwardslash" paths on your clipboard

```{r}

#helpful environment modifications

  #(optional) set number of columns to display in "View()" function to 1000 - revert to 50 if you don't like the behavior
  if(F) rstudioapi::writeRStudioPreference("data_viewer_max_columns", 1000L)

  #turn off scientific notation 
  options(scipen = 10E6)
  
  #prevent strings being read as factors
  options(stringsAsFactors = F)

#helpful functions
  
  #convenience function for package loading / update /  install
    library1 = function(pkgs=NA,load=T,update=F){
      if(!is.na(pkgs[1])){
        #parse existing packages
        old.pkgs <- pkgs[(pkgs %in% installed.packages()[, "Package"])]      
        new.pkgs <- pkgs[!(pkgs %in% installed.packages()[, "Package"])]
        #update and install
        if( (length(old.pkgs)>0) & update) update.packages(old.pkgs, dependencies = TRUE)  
        if(length(new.pkgs)>0) install.packages(new.pkgs, dependencies = TRUE)
        if(load) sapply(pkgs, function(...)try(require(...)), character.only = TRUE)
      }
    }

  #convenience function for more usable version of writeClipboard - overcomes newline issues
    writeClipboard2 = function(x) cat(x, file="clipboard")
  
  #generic file path editing function - forwards or backwards
    reslash<- function(x=NA, slash=c("back","forward")){
      if(is.na(x)) x = readClipboard()
      if(slash[1]=="back") path <- shQuote(gsub("/","\\\\",gsub("\\", "\\\\", x, fixed = TRUE), fixed = TRUE))
      if(slash[1]=="forward") path <- shQuote(gsub("\\", "/", x, fixed = TRUE))
      writeClipboard2(path)
      return(path)
    }
  
  #convert commas into newline commas
  nl1 = function(x=NA,newline="\n",npad=4, space_remove=T, newline_remove=T){
     #grab from clipboard
      if(is.na(x)) xin = paste(readClipboard(),collapse="") else xin = x
     #remove spaces and newlines  
      if (space_remove)   xin <- gsub(" ", "", xin, fixed = TRUE)
      if (newline_remove) xin <- gsub("\\R", "", xin, perl = TRUE)

     #generate generic padding and padded newline
      pad_in = strrep(" ", npad)
      newline_pad = paste0(newline,pad_in)
     #add padding and newlines to original strin  
      xin = paste0(" ",pad_in,gsub(",",paste0(newline_pad,","),xin))
      #xin = gsub(",",paste0(newline_pad,","),xin)
     #write / return
      writeClipboard2(xin)
      cat("--- Result copied to clipboard ---\n")
      cat(xin, "\n")
      return(invisible(xin))
  }
  makeActiveBinding("nl", nl1, .GlobalEnv)

  #correctly escape back slashes and quote path - a parameterized version o f   
  bs1=function(x=NA){reslash(x,slash="back")}
  makeActiveBinding("bs", bs1, .GlobalEnv)
    
  #correctly escape back slashes and quote path   
  fs1=function(x=NA){reslash(x,slash="forward")}
  makeActiveBinding("fs", fs1, .GlobalEnv)
  
  #function to grab names from data.frame, quote them, place commas between them, and send to clipboard
  # - version 2 adds newline at 50 chars
    nmsVec=function(x,char_break=70,newline="\n"){
      nms_in = names(x)
      vec_i = paste0("'",nms_in[1],"'")
      char_break_in = char_break
      for(i in 2:length(nms_in)){
        vec_ln = nchar(vec_i)
        if((vec_ln + nchar(nms_in[i])) >= char_break_in){
          char_break_in = char_break + nchar(vec_i) 
          vec_i = paste0(vec_i,"\n    ,'",nms_in[i],"'")
        }else{
          vec_i = paste0(vec_i,",'",nms_in[i],"'")
        }
      }
      #add parentheses  
      vec_i = paste("c(",vec_i,"  \n    )",sep="",collapse="")
      #write / return
      writeClipboard2(vec_i)
      cat(vec_i)
      return(vec_i)
    }

  #common stats functions
    mse = function(err,na.rm=T){ mean(err^2,na.rm=na.rm) } 
    rmse = function(err,na.rm=T){ sqrt(mean(err^2,na.rm=na.rm)) }   
    rsq = function(err, y){ (1 - sum(err^2) / sum((y-mean(y))^2) )  }
    cv = function(x,fun=median,...){sd(x,...)/fun(x,...)}
    cvpct = function(x,fun=median,mult=100,...){mult*sd(x,...)/fun(x,...)}
    #standard se function
    se = function(err,na.rm=T){ sqrt(mean(err^2,na.rm=na.rm)) / sqrt(length(err)) }   
    #se when n is known and need to add zeros
    se_nzero = function(err, n,na.rm=T){
      n_err = length(err)
      n_zero = n - n_err
      if(n_zero>0) err_in = c(rep(0,n_zero),err)
      else err_in = err
      sqrt(mean(err_in^2,na.rm=na.rm)) / sqrt(n) 
    } 
    
```

## load some common libraries
 - use custom library1 function defined above that installs / loads packages
```{r}

  #install.packages("mltools",type="source")
  #remotes::install_github("jstrunk001/RSForTools")
  #remotes::install_github("jstrunk001/RForInvt")

  library1(c(
     # "chattr",
       "data.table"
     #,"devtools"
	   #,"dplyr"
     #,"DBI"
     #,"exactextractr"
     #,"future"
      ,"gridExtra"
      ,"ggplot2"   
    # ,"ggmap"
	  # ,"gganimate"
	  # ,"ggpmisc	"
	  # ,"gtsummary"
    # ,"kableExtra"
    # ,"lidR"
  	# ,"leaps"
  	# ,"lme4"
  	# ,"MixRF"
    # ,"odbc"
    # ,"openxlsx"
    # ,"parsnip"
	    ,"parallel"
    # ,"ranger"
    # ,"reactable"
    # ,"reactablefmtr"
    # ,"readxl"
    # ,"RForInvt"
    # ,"RSForTools"
    # ,"RSQLite"
    # ,"rstac"
      ,"sf"
  	# ,"survey" 
    # ,"terra"
    # ,"tidymodels"
	  # ,"tmap"	
    # ,"tune"	
    # ,"vip"
    )
    ,load=T, update=F
    )

  #if(isNamespaceLoaded("ggmap")) ggmap::register_google(key = "???")

```


## load data - some examples of ways to load various data types. 

```{r}

#examples
if(F){ 
	
  library(RSQLite)
  library(DBI)
	
	#Example make some fake data and write to versioned files 
  
  	#create new output file paths for this version - increment = T  
    dir_temp = "c:/temp"
  	file_csv = RForInvt::file_increment(file.path(dir_temp,"some_data.csv"), increment=T)
  	file_rds = RForInvt::file_increment(file.path(dir_temp,"some_data.RDS"), increment=T)
  	file_db = RForInvt::file_increment(file.path(dir_temp,"some_data.sqlite"), increment=T)
  	
  	#simulate / write small simulated table of data
  	
  		#simulate
  		test_dat = data.frame(plot=1:5,some_y=rnorm(5))
  		
  		#write csv
  		write.csv(test_dat,  file_csv )
  		
  		#save R binary format
  		saveRDS(test_dat, file_rds)
  		
  		#save to database
  	  con_this = dbConnect(RSQLite::SQLite(), file_db )
  	  	dbWriteTable( con_this , "pldat" , test_dat , overwrite = T )
  	  dbDisconnect(con_this)  
	
	#grab latest versions of data - increment = F
	  
	  #prep file names
    dir_temp = "c:/temp"
  	file_csv = RForInvt::file_increment(file.path(dir_temp,"some_data.csv"), increment=F)
  	file_rds = RForInvt::file_increment(file.path(dir_temp,"some_data.RDS"), increment=F)
  	file_db = RForInvt::file_increment(file.path(dir_temp,"some_data.sqlite"), increment=F)
  	
  	#read data csv/rds
  	dat_csv = read.csv(file_csv)
  	dat_rds = readRDS(file_rds)
  	
  	#read sqlite data
    con_this = dbConnect(RSQLite::SQLite(), file_db )
    	dat_db = dbReadTable(con_this ,"pldat" )
    dbDisconnect(con_this)  
    
    #add sf::st_write example
    
    #add terra::writeRaster example
    
}


```

## example of how to save versions figures to tiff and pdf

```{r, eval=FALSE}

	#create versioned files names
  dir_temp = "c:/temp"
	file_tif = RForInvt::file_increment(file.path(dir_temp,"some_plots.tif"), increment=T)
	file_pdf = RForInvt::file_increment(file.path(dir_temp,"some_plots.pdf"), increment=T)

	#simulate / write small simulated table of data
	
		#simulate
		test_dat1 = data.frame( plot=1:5,y=rnorm(5), x=rnorm(5) )
		
		#save tif - show how to combine exponent with regular text
		tiff(file_tif , width = 6, height = 4, units = "in", res = 600)
		  plot(y~x, data = test_dat1 , main = "test_dat1", xlab=expression(paste("bigfoot height ",cm^2)), ylab="ft")
		dev.off()
		
		#save pdf - show how to save 2 figures in same file
		pdf(file_pdf , width = 6, height = 4, units = "in", res = 600)
		  plot(y~x, data = test_dat1 , main = "test_dat1", xlab=expression(paste("bigfoot height ",cm^2)), ylab="ft")
		  plot(y~x, data = test_dat1 , main = "test_dat1", xlab=expression(paste("bigfoot height ",cm^2)), ylab="ft")
		dev.off()



```

## 

```{r, eval=FALSE}

```

## 

```{r, eval=FALSE}

```

## 

```{r, eval=FALSE}

```


## 

```{r, eval=FALSE}

```

## 

```{r, eval=FALSE}

```

## 

```{r, eval=FALSE}

```
