---
title: "Untitled"
author: "Jacob Strunk"
date: "8/5/2021"
output: html_document
---
## This is an alternative document template for RStudio 
 Simply drop this file in the templates folder for Rstudio and replace the existing document.RMD
 Feel free to make any other changes, like placing helpful functions in the first chunk or 
 preparing some canned code for reading data, connecting to databases, etc. or simply remove 
 what is here 


```{r, environment}

#turn off scientific notation and prevent strings being read as factors
options(scipen = 10E6, stringsAsFactors = F)

```


## helpful functions and some options setup

```{r, helpers}

writeClipboard2 = function(x) writeClipboard(charToRaw(paste0(x, ' ')))
reslash<- function(x=NA, slash=c("back","forward")){
  if(is.na(x)) x = readClipboard()
  if(slash[1]=="back") path <- shQuote(gsub("/","\\\\",gsub("\\", "\\\\", x, fixed = TRUE), fixed = TRUE))
  if(slash[1]=="forward") path <- shQuote(gsub("\\", "/", x, fixed = TRUE))
  writeClipboard2(path)
  return(path)
}
bs=function(x=NA){reslash(x,slash="back")}
fs=function(x=NA){reslash(x,slash="forward")}
nmsVec=function(x){x=paste("c('",paste(names(x),collapse="','"),"')",sep=""); writeClipboard2(x);return(x)}


```

## load data

```{r}




```

## 

```{r, eval=FALSE}

```

## 

```{r, eval=FALSE}

```

## 

```{r, eval=FALSE}

```

## 

```{r, eval=FALSE}

```


## 

```{r, eval=FALSE}

```

## 

```{r, eval=FALSE}

```

## 

```{r, eval=FALSE}

```