---
title: "County-level FIA estimation in western Washington"
author: "RForInvt"
date: "`r format(Sys.Date())`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE)
```

## What this shows

A complete path from a raw FIA database (the FIA DataMart SQLite) to
**design-based, EVALIDator-style estimates** for western Washington counties,
and then a side-by-side comparison of **design-based vs model-based annualized
estimation**. The pipeline is the same four steps every time:

```
fia_trees() + fia_plots()  ->  fia_compile_trees()  ->  fia_compile_plots()  ->  fia_estimate()
   (data layer)                (per-tree attrs)         (per-acre, zero-filled)   (post-stratified)
```

Everything in the **runnable** chunks below uses the small synthetic fixture
bundled with the package (`FIADB_demo.db`: STATECD 53, the westside timber
counties King/Lewis/Grays Harbor, PNW species, a 2019 inventory remeasured in
2021). It is schema-faithful to the DataMart so the code is identical to what
you would run on a real state download — but the **numbers are synthetic and do
not match EVALIDator**. The `eval=FALSE` chunks near the end show how to point
the same code at a real western-WA SQLite and cross-check against EVALIDator.

```{r load}
library(RForInvt)
library(DBI); library(RSQLite)

db_path <- system.file("extdata", "FIADB_demo.db", package = "RForInvt")
db <- fia_db(db_path)
db

# evaluations available in this database
fia_evalid(db, statecd = 53, eval_type = "EXPVOL")
```

## 1. Single-year county estimates

Pick the most recent volume evaluation, assemble the tree and plot frames,
compile, and estimate net cubic-foot volume per forest acre and in total, by
county.

```{r single-year}
ev <- fia_evalid(db, statecd = 53, eval_type = "EXPVOL", most_recent = TRUE)

tr <- fia_trees(db, evalid = ev)          # live trees, with per-acre TPA_EXP
pl <- fia_plots(db, evalid = ev)          # condition rows + design columns

trc <- fia_compile_trees(tr, vol_source = "fiadb")   # ba_ft, dbcl, vol_cf_net, ...
plc <- fia_compile_plots(trc, pl)                    # per-acre, nonforest plots zero-filled

# volume per forest acre, and total, by county
fia_estimate(plc, vars = "vol_cf_net", by = "COUNTYCD", type = "per_acre")
fia_estimate(plc, vars = "vol_cf_net", by = "COUNTYCD", type = "total")

# forest area by county (the ratio denominator)
fia_estimate(plc, by = "COUNTYCD", type = "area")
```

`fia_estimate()` reports the estimate, its sampling variance, standard error,
`se_pct`, and the plot counts, following the Bechtold & Patterson (2005)
post-stratified estimator. The total is identically the EXPNS-weighted sum of
the per-acre plot values; per-acre is the ratio of the volume total to the
forest-area total.

## 2. Multiple years (a trend)

`fia_estimate_annual()` runs the whole pipeline across several annual
evaluations and stacks the result into a long trend frame.

```{r multi-year}
trend <- fia_estimate_annual(db, statecd = 53, years = c(2019, 2021),
                             vars = "vol_cf_net", by = "COUNTYCD",
                             type = "per_acre")
trend[, c("year", "COUNTYCD", "estimate", "se")]
```

```{r trend-plot, eval=FALSE}
# a simple trend plot by county (base graphics)
cts <- sort(unique(trend$COUNTYCD))
plot(NA, xlim = range(trend$year), ylim = range(trend$estimate),
     xlab = "year", ylab = "net cu-ft / forest acre")
for (i in seq_along(cts)) {
  d <- trend[trend$COUNTYCD == cts[i], ]
  lines(d$year, d$estimate, type = "b", col = i)
}
legend("topleft", legend = cts, col = seq_along(cts), lty = 1, title = "county")
```

In real data these years are FIA *panels*; remember that consecutive EVALIDs are
moving windows that share most plots, so a "trend" of overlapping evaluations is
smoother (and more correlated) than a trend of independent annual panels.

## 3. Change between two evaluations

`fia_estimate_change(mode = "remeas")` pairs each current plot to its previous
measurement (via `PREV_PLT_CN`), differences the per-acre attribute on the
paired plots, and estimates the change with the current design — so the variance
is a proper paired-plot variance, not the difference of two independent SEs.

```{r change}
chg <- fia_estimate_change(db, statecd = 53,
                           evalid_t1 = 531901, evalid_t2 = 532101,
                           mode = "remeas", vars = "vol_cf_net",
                           by = "COUNTYCD", type = "per_acre")
chg[, c("COUNTYCD", "estimate", "se")]
```

For a components view of change (net growth, removals, mortality), FIA ships
growth-removal-mortality evaluations (`EXPGROW`/`EXPREMV`/`EXPMORT`) and the
`TREE_GRM_COMPONENT` table. `mode = "grm"` is the place that will live; it is not
yet implemented (it stops with a message).

## 4. Components: volume by species and diameter class

Recompile the trees with the species / diameter-class spreading functions so
each tree carries wide component columns, then estimate every component column
and tidy the result back to long form.

```{r components}
# species breakdown of net cubic volume
trc_sp <- fia_compile_trees(tr, vol_source = "fiadb",
                            fns_compute = list(ba_ft, dbcl, fia_vol, spp_y),
                            vars_group  = c("vol_cf_net"))
plc_sp <- fia_compile_plots(trc_sp, pl)

comp <- fia_estimate(plc_sp, vars = "^vol_cf_net_SPCD_", by = "COUNTYCD",
                     type = "total")
comp <- fia_components_long(comp)
head(comp[, c("COUNTYCD", "SPCD", "estimate", "se")])

# components sum to the overall total (a useful check)
agg <- aggregate(estimate ~ COUNTYCD, data = comp, FUN = sum)
whole <- fia_estimate(plc_sp, vars = "vol_cf_net", by = "COUNTYCD", type = "total")
merge(agg, whole[, c("COUNTYCD", "estimate")], by = "COUNTYCD",
      suffixes = c("_components", "_whole"))
```

For a two-way species-by-diameter-class breakdown, add `dbcl_spp_y` to
`fns_compute` and estimate `vars = "^vol_cf_net_SPCD_.*_dbcl_"`;
`fia_components_long()` parses both the `SPCD` and `dbcl` out of the column name.

## 5. Annualized estimation: design-based vs model-based

This is the heart of the western-WA workflow. For a single target year, only a
fraction of plots are measured (~1/10 in the West), so a design-based estimate
for that year alone is noisy. Four strategies trade bias for precision:

1. **Design-based, single panel** — only plots measured that year. Unbiased,
   noisy. `fia_annual_panel()` selects the panel; `fia_estimate()` estimates it.
2. **Model-assisted (GREG)** — fit a growth model on remeasured plots, predict
   *all* plots forward to the target year, then bias-correct with the
   design-weighted residuals on the measured plots. The residuals are put on a
   common annual footing first (`fia_annualize_resid()`):
   `e_annual = e_T / T`, then `e_k = (Y - t1) * e_annual`.
   `fia_estimate_greg()` combines the two. Lower variance, ~design-consistent.
3. **Pure model-based** — the model predictions for every plot, no correction
   (`fia_estimate_model()`); characterize bias/precision by cross-validation
   (`fia_growth_cv()`). Lowest variance, exposed to model bias.
4. **FVS projection** — grow each plot's previous measurement to the target year
   with FVS (`fia_fvs_*`), then compile and estimate through the same pipeline.
   Process-based; carries FVS calibration bias.

```{r annualized-setup}
# current (2021) and previous (2019) compiled plot frames
cur  <- fia_compile_plots(fia_compile_trees(fia_trees(db, evalid = 532101)),
                          fia_plots(db, evalid = 532101))
prev <- fia_compile_plots(fia_compile_trees(fia_trees(db, evalid = 531901)),
                          fia_plots(db, evalid = 531901))

# pair current plots to their previous measurement; build predictors + interval
pr <- prev[, c("PLT_CN", "vol_cf_net", "ba_ft", "stems")]
names(pr) <- c("PREV_PLT_CN", "vol_t1", "ba_t1", "stems_t1")
pairs <- merge(cur, pr, by = "PREV_PLT_CN")
pairs$prev_yr  <- 2019
pairs$curr_yr  <- pairs$MEASYEAR
pairs$interval <- pairs$curr_yr - pairs$prev_yr
```

```{r annualized-model}
# fit a growth/projection model on the remeasured plots
m <- fia_growth_model(pairs, response = "vol_cf_net",
                      predictors = c("vol_t1", "ba_t1", "stems_t1", "interval"))
m

# cross-validated bias / RMSE / R^2 (approach-3 precision & bias)
fia_growth_cv(pairs, response = "vol_cf_net",
              predictors = c("vol_t1", "ba_t1", "stems_t1", "interval"),
              k = 5, seed = 1)

# predictions at the target year + annualized residuals
target_year <- 2021
pairs <- fia_annualize_resid(m, pairs, response = "vol_cf_net",
                             prev_year_col = "prev_yr", curr_year_col = "curr_yr",
                             target_year = target_year)
head(pairs[, c("PLT_CN", "vol_cf_net", ".pred", ".resid_periodic", ".resid_target")])
```

```{r annualized-compare}
# one statewide domain so each approach has all plots to work with
pairs$ALL <- "wWA"

# (ref) design-based using ALL measured plots -- the full-measurement reference
ref <- fia_estimate(pairs, vars = "vol_cf_net", by = "ALL", type = "total")

# (3) pure model-based: predictions for every plot
mod <- fia_estimate_model(pairs, ".pred", by = "ALL", type = "total")

# (2) GREG: predictions everywhere + design-weighted annualized residual
pairs$.resid_obs <- pairs$vol_cf_net - pairs$.pred   # full residual here (demo)
greg <- fia_estimate_greg(pairs, pred_col = ".pred", df_sample = pairs,
                          resid_col = ".resid_obs", by = "ALL", type = "total")

compare <- data.frame(
  approach = c("design (all plots)", "model-based", "GREG"),
  estimate = c(ref$estimate, mod$estimate, greg$estimate),
  se       = c(ref$se,       mod$se,       greg$se)
)
compare
```

A caveat specific to this fixture: it is a **full remeasurement** (every plot
measured in both years), not a true rotating panel, so the approaches largely
agree and approach 1 here uses all plots. The headline trade-off — approach 1
*noisy*, approach 3 *precise but possibly biased*, approach 2 *precise and
bias-corrected* — only becomes visible when a single year is ~1/10 of the plots
and each annual panel carries its own EXPNS. That is what the real-data section
demonstrates.

## 6. Pointing at real western-WA data (and checking EVALIDator)

The code is identical; only the database changes. Download the Washington
SQLite from the FIA DataMart
(<https://apps.fs.usda.gov/fia/datamart/datamart.html>), then:

```{r real-wwa, eval=FALSE}
db_wa <- fia_db("SQLite_FIADB_WA.db")               # ~1 GB state download

ev <- fia_evalid(db_wa, statecd = 53, eval_type = "EXPVOL", most_recent = TRUE)
tr <- fia_trees(db_wa, evalid = ev)
pl <- fia_plots(db_wa, evalid = ev,
                vars_keep = c("SLOPE", "ASPECT", "ELEV"))  # site covariates for modeling
plc <- fia_compile_plots(fia_compile_trees(tr), pl)

# statewide and by-county net cubic volume -- compare to EVALIDator totals
fia_estimate(plc, vars = "vol_cf_net", by = "COUNTYCD", type = "total")

# EVALIDator: pick "Net cubic-foot volume of live trees (>= 5 in)" for the same
# EVALID and confirm the per-county totals agree within rounding.
```

```{r wwa-map, eval=FALSE}
# a county choropleth of volume per acre (needs a county polygon layer)
library(sf)
est  <- fia_estimate(plc, vars = "vol_cf_net", by = "COUNTYCD", type = "per_acre")
cnty <- sf::st_read("wa_counties.shp")               # your county polygons
cnty <- merge(cnty, est, by.x = "COUNTYFP", by.y = "COUNTYCD")
plot(cnty["estimate"], main = "Net cu-ft / forest acre")
```

For the annualized comparison on real data, build the panel explicitly with
`fia_annual_panel(plc, year = 2017)` (one measured panel), fit
`fia_growth_model()` on remeasured plots from earlier evaluations, predict the
full plot set at 2017, and run approaches 1-3 exactly as above. Because each
single-year panel has its own EXPNS, the design-only SE (approach 1) will be
visibly larger than the GREG and model-based SEs.

## 7. Optional: NVEL volumes and FVS projection

Swap `vol_source = "nvel"` to recompute volumes with the National Volume
Estimator Library instead of using the FIADB columns (useful for custom merch
specs or consistency with a non-FIA inventory). This requires the bundled vollib
DLL.

```{r nvel, eval=FALSE}
trc_nvel <- fia_compile_trees(tr, vol_source = "nvel", region = 6)
plc_nvel <- fia_compile_plots(trc_nvel, pl)
fia_estimate(plc_nvel, vars = "vol_cf_net", by = "COUNTYCD", type = "per_acre")
```

And the FVS bridge projects plots forward, then re-enters the same compile +
estimate pipeline (requires an installed FVS variant executable):

```{r fvs, eval=FALSE}
db_fvs_in  <- fia_fvs_input(db_wa, evalid = ev, dir_out = "c:/temp/fia_fvs",
                            variant = "PN")
db_fvs_out <- fia_fvs_run(db_fvs_in, fvs_path = "C:/FVSbin/FVSpn.exe",
                          numcycle = 1, timeint = 10)
proj <- fia_fvs_compile(db_fvs_out, year = 2031)
# proj flows straight into fia_compile_plots() + fia_estimate() for a projected
# county estimate -- "current" and "projected" estimates are symmetric.
```
