---
title: "Model Archive: NBEL biomass & NVEL volume"
author: "Jacob Strunk"
date: "2026-06-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(scipen = 10e6, stringsAsFactors = FALSE)
```

This example shows how the shipped model archive carries USFS **National Biomass
Estimator Library (NBEL)** equations as `closed_form` models and **National
Volume Estimator Library (NVEL)** equations as `nvel` models, and how to
discover and apply them with `model_archive_list()`, `model_archive_load()`, and
`model_predict()`.

```{r library}
library(RForInvt)
library(data.table)
```

## 1. Catalog the shipped archive

`model_archive_list()` reads the shipped registry (`extdata/models.csv`) without
loading any sidecars. Each row is one model; biomass components are named
`bm_<component>` (e.g. `bm_stw` = stem wood, `bm_fot` = foliage total).

```{r list}
cat_all = model_archive_list(package = "RForInvt")

# how many models, by region and class?
cat_all[, .N, by = .(region_code, model_class)]

# biomass components available for loblolly pine
cat_all[spp == "pinus.taeda" & model_class == "closed_form",
        .(model_id, response, n, r2_pd)]
```

Filter to a region / species / component:

```{r filter}
se_doug = model_archive_list(
  filter  = quote(region_code == "se" & spp == "pinus.taeda"),
  package = "RForInvt"
)
se_doug[, .(model_id, response, formula = substr(formula, 1, 40))][1:5]
```

## 2. Predict NBEL biomass on a tree list

Build a small tree list, then apply one or more biomass models. Map your data
columns to the canonical predictor names (`dbh` in inches, `ht` in feet) via
`nms_dat_predict`. Each model appends one column named by its `response`.

```{r biomass-predict}
trees = data.table(
  tree = 1:4,
  DBH  = c(6, 10, 14, 18),   # inches
  HT   = c(40, 60, 78, 92)   # feet
)

# load a few loblolly biomass components
lobl = model_archive_load(
  filter  = quote(spp == "pinus.taeda" & model_class == "closed_form"),
  package = "RForInvt"
)
params = data.table::rbindlist(lapply(lobl, `[[`, "row"), fill = TRUE)

out = model_predict(
  dat             = trees,
  nms_dat_predict = c(dbh = "DBH", ht = "HT"),
  params          = params,
  package         = "RForInvt"
)
out
```

Predicted biomass is in **pounds** for each component. The coefficient
convention (published `a..e` in `b0..b4`; unit factors in `b5..b7`; log-bias in
`b8`) is documented in `inst/docs/model_archive_spec.md`.

## 3. Predict NVEL volume

NVEL volume equations ride in the same archive as `model_class == "nvel"`. They
are DLL-backed (the bundled `vollib`), so this section runs only where the DLL
is available.

```{r nvel-predict}
nvel_ok = RForInvt:::.nvel_dll_available()

if (nvel_ok) {
  vol_models = model_archive_load(filter = quote(model_class == "nvel"),
                                  package = "RForInvt")
  vrow = vol_models[[1]]$row
  vol = model_predict(
    dat             = trees,
    nms_dat_predict = c(dbh = "DBH", ht = "HT"),
    params          = vrow,
    package         = "RForInvt"
  )
  vol[, .(tree, DBH, HT, vol_tcf)]
} else {
  message("NVEL vollib DLL not available; skipping volume prediction.")
}
```

You can register your own NVEL equation into a writable archive with
`nvel_volume_register()`:

```{r nvel-register, eval=FALSE}
arch = file.path(tempdir(), "my_archive")
nvel_volume_register(archive_dir = arch, spcd = 202, region = 6,
                     spp = "pseudotsuga.menziesii", region_code = "pnw")
```

## 4. Add your own model

Any model — a fitted `randomForest`/`lme4`/`xgboost` object, or a closed-form
equation — can be saved with `model_archive_save()` into a writable
`archive_dir`, then loaded and predicted the same way. See
`?model_archive_save` and `inst/docs/model_archive_spec.md`.
